Analysing the three-dimensional structure of P. falciparum

### 2017-07-08 doc/manuscript/paper.tex
Processing lemieux's data

While these approaches are widely used, since better motivated and more
flexible models to infer 3D consensus structures from contact counts have been
proposed.

@@

The relationship between genomic distance $s$ and $c$ can be derived from the
data, by fitting a linear model between the log expected count and the genomic
distances.

@@

Note that in practice, the constraints are fulfilled naturally by the data,
and thus can be dropped. In addition of drastically reducing the computation
time, this allows to use generic publicly available software such as chromSDE
(which implements a variant of the model defined here), or Pastis-MDS (which
implements exactly the unconstrained model defined here).

\subsubsection{Modeling contact counts as a Poisson distribution}

\subsubsection{Structure stability}
- Structure stability
- at least 100 iterations
- maybe compare results with chromSDE, PCA \& ShRec3D 
-

\subsubsection{Volume exclusion models}
blha

>>>>

@@

@@

The relationship between genomic distance $s$ and $c$ can be derived from the
data, by fitting a linear model between the log expected count and the genomic
distances and extracting the slope coefficient $\alpha$. We thus obtain, up to
a scaling factor $\beta$ a function to map contact counts to wish distances:

\begin{equation}
\label{eq:count-to-distance}
\delta_{ij} = \beta c^{\alpha/3}
\end{equation}

The scaling factor depends both on the sequencing depth of the libraries, but
also on the nucleus size of the organims. \citet{ay:three-dimensional} infers
3D models for a range of parameters and keeps the models that are the most
plausible, under the assumption that beads should be as uniformely distributed
in the nucleus as possible. 0-contact counts can either be ignored, or mapped
to the largest distance possible. Note that for high coverage data set, the
amount of non-interacting beads is sufficiently small that including those or
not in the optimization will make very little difference.

\subsubsection{Reproducing the results using a simpler model}

Two simplifications can be brought to the method proposed by
\citet{ay:three-dimensional}.

First, the ``count-to-distance'' relationship is based on two steps: (1) find
the relationship between genomic distances and counts; (2) from the
coefficient, deduce what type of polymer the DNA of this particular organism
is closest two (fractal globule or equilibrium); (3) use the relationship
between genomic distance $s$ and euclidean distance $d$ for that particular
type of polymer. Yet, it is likely that the probability of contact count
between two elements depends mostly on the average distance separating the two
beads. As noted in \citet{varoquaux:statistical}, while the relationship
between both genomic distance $s$ and expected contact count $c$ and the
relationship between genomic distance $s$ and euclidean distance $c$,
combining both yields the same count-to-distance function for both polymer
type:

\begin{equation}
\delta_{ij} = \beta c^{-1/3}
\end{equation}

While the difference may seem subtile, fitting the linear model to find
$\alpha$ is not as straightforward as it seems, as the range of genomic
distance used can have a large impact on the final coefficient.

The second simplification is to perform an unconstraint optimization. Indeed,
in practice, the constraints are fulfilled naturally by the data and thus can
be dropped from the optimization. In addition of drastically reducing the
computation time, this allows to use generic publicly available software such
as chromSDE (which implements a variant of the model defined here), or
Pastis-MDS (which implements exactly the unconstrained model defined here).

\subsubsection{Structure stability}

Both the constrained and unconstained problem stated above are not convex. The
reader may well ask how sensitive the result is to initialize. In short, it is
not. To study the stability of the results with respect to the initialization,
I perform for each stage presented here 1000 optimization, starting from
different initialization.

The first question that arise is simply: are structures from the same time
points but from a different start ponit more alike than structures from
different time points? To answer this question, I compute for each structure
and each time point a feature matrix, composed of a subsampled pairwise
euclidean distances of each structure. The reason for subsampling is simply so
that the resulting feature matrix fits in memory on a laptop. I then compute a
stochastic PCA on this feature matrix and plot the projection in 2D space
using the 3 first components. The results clearly show that structures from
the same time point cluster strongly. Repeating this experiment on singly
chromosomes yields the same result.

The second question one can ask is how do the structures differ. Tackling this
question is very challenging, but can be reformulated in a much easier way.
Are the hallmarks of interest conserved across structures of a same stage?
\citet{ay:three-dimensional} and \citet{lemieux:genome-wide} both identified
{\em P. falciparum} folded in very specific ways, with VRSSM genes highly
interacting. \citet{ay:three-dimensional} also observed strong clustering of
the centromeres, and enrichment in interaction at the telomeres. These
observation can lead to a rigourous approach at identifying whether families
of loci clustered in the structures and even quantifying this in a rigourous
manner.

\subsubsection{Volume exclusion models}
